Notes on Design and Implementation of Anglican

### 2015-07-24 paper/paper.tex
white paper in progress --- transformation

>>>>

Compilation of Anglican into Clojure relies on the Clojure
\textit{macro} facility. However, the compilation algorithm is
implemented as a library of \textit{functions} in namespace
\texttt{anglican.trap}, which are invoked by macros. Namespace
\texttt{anglican.trap-test} contains many transformation tests,
which both help assure proper functioning of the transformation
functions, and serve as illustrative examples of individual
transformations, such as of functions, flow control, and
probabilistic contstructs.

The CPS transformation is organized in top-down manner. The
top-level function is  \texttt{cps-of-expression}, which
receives an expression and a continuation, and returns the
expression in the CPS form, with the computed result passed to
the continuation. A continuation accepts \texttt{two arguments}:
\begin{itemize}
	\item the computed value;
	\item the internal state, bound to the local variable
		\texttt{\$state} in every lexical scope.
\end{itemize}
The state (\texttt{\$state}) is threaded through the computation
and contains data used by inference algorithm. \texttt{\$state}
is a Clojure hash map, and the map entries are
algorithm-dependent. Except for transformation of
\texttt{mem}, the memoization form, CPS transformation routines
are not aware of contents of \texttt{\$state}, do not access or
modify it directly, but rather just thread the state unmodified
through the computation. Algorithm-specific handlers of
checkpoints corresponding to the probabilistic forms
(\texttt{sample}, \texttt{observe}) modify the
state and reinject a new state into the computation.

\subsection{Expression Kinds}

There are three different kinds of inputs to CPS transformation:
\begin{itemize}
	\item literals, which are passed as an argument to the
		continuation unmodified;
	\item value expressions (e.g. the \texttt{fn} form) 
		(called \textit{opaque}
		expressions in the code) which must be transformed to
		CPS, but the transformed object is passed to the
		continuation as a whole, opaquely;
	\item general expressions (let's call them
		\textit{transparent}), through which the continuation is
		threaded in an expression-specific way, and can be
		called in multiple locations of the CPS-transformed
		code, such as in all branches of an \texttt{if}
		statement.
\end{itemize}

\subsection{Literals}

Literals are the same in Anglican and Clojure. They are 
left unmodified. Literals are a subset of opaque expressions,
and are identified by test \texttt{simple?} called from
\texttt{opaque?}.  However, The Clojure syntax has a peculiarity
of using the syntax of compound literals (vectors, hash maps,
and sets) for data constructors. Hence, compound literals must
be traversed recursively, and if there is a nested non-literal
component, transformed into a call to the corresponding data
constructor. Functions \texttt{cps-of-vector},
\texttt{cps-of-hash-map}, \texttt{cps-of-set}, called from
\texttt{cps-of-expression}, transform Clojure constructor syntax
(\texttt{[...]}, \texttt{\{...\}}, \texttt{\#\{...\}}) into the
corresponding calls.

\subsection{Opaque Expressions}

Opaque, or value, expressions, have a different shape in the
original and the CPS form. However, their CPS form follows the
pattern \texttt{(continuation transformed-expression)}, and thus
the transformation does not depend on the continuation, and
can be accomplished without passing the continuation as a
transformation argument. Primitive (non-CPS) procedures used in
Anglican code, \texttt{(fn ...)} forms, and \text{(mem ...)}
forms are opaque and transformed by
\texttt{primitive-procedure-cps}, \texttt{fn-cps}, and
\texttt{mem-cps}, correspondingly.

\subsection{General Expressions}
